---
title: "Assignment 3: Data Exploration"
author: "Quinn Bankson"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,  message = FALSE}
library(tidyverse)
library(lubridate) 

Neonics <- read.csv("/Users/mac/Documents/EDE_Fall2023/Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("/Users/mac/Documents/EDE_Fall2023/Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)


```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids are extremely common instectidices. Although they are effective at limitting targetted populations of instects, neonictinoids can have effects on insect populations outside of pests that are targetted. It is important to show the effect of neonictinoids on both target populations of insects and on populations like pollinators. One may use this data to judge if neonictinoids are worth it in terms of reducing pests versus reducing desirable insect populations. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Debris on forest floors are essential for determining flooding outcomes, fire risk, nutrient cycles, and even carbon sequesteration possibilities. Keeping track of the debris on forest floors and how long it takes for the ecosystem to absorb debris back into the soil can give researchers time frames for seasonal fire risks and flood risks. The presence of these debris also can inform researchers about the quality and nutrient make up of soil. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Takes place in 20 40 x 40 meter plots.
 2. Trap placement for collection within plots may either be randomized or targeted depending on vegetation cover. 
 3. Ground traps are sampled only once a year, and elevated traps are sampled more frequently (every month or so for evergreen areas and every 2 weeks for deciduous areas). 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r, results = TRUE}
dim(Neonics)
summary(Neonics)
head(Neonics)

#Dimensions: 4623 rows for 30 columns (variables). 
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer: Some of the most common effects are population, mortality, and behavior. These are of interest because they are critical to maintaining an ecosystem. Rapid increases in mortality, decreases in population, or change in behavior for key organisms in an ecosystem can cause a collapse. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
sort(summary(Neonics$Species.Common.Name))

```

> Answer: The most commonly studied species include Honey Bees, Parasitic Wasps, Buf Tailed Bumblebees, Asian Lady Beetles, Euonymus Scale, and Wireworms. The insects have things in common. The bees are common pollinator and are critical for ecosystem health. Parasitic wasps and and Asian Lady Beetles are known for consuming pests as they are predators Scale and Wire worm are common pests that can damage crops. The interaction between the populations of predators, pollinators, and pests is an important aspect for studies. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.)

```

> Answer: The column represents a factor class of data. It categorizes data rather tahn representing only the discrete numeric value. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r fig.height = 3, fig.width = 4}
ggplot(Neonics, aes(x = Publication.Year)) +
  geom_freqpoly()
``` 

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r fig.height = 3, fig.width = 4}
ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  geom_freqpoly()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Lab locations are the most frequent locations for studies after 2010. The second most common test location is "field natural". "Field Natural" was the most frequent test location from 1990-2000, and briefly in 2007-2010. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer:

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)

Litter$collectDate <- ymd(Litter$collectDate)

class(Litter$collectDate)

unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)

```

> Answer: Unique displayes the data for each of the distinct values in the plotID column. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter, aes(x=functionalGroup)) +
  geom_bar(fill = "darkorchid") + 
  labs(title="Type of Litter Collected at Niwot Ridge", x="Litter Type") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}
ggplot(Litter) + 
  geom_boxplot(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) + 
  geom_violin(aes(x = functionalGroup, y = dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
